Transcriptomic Analysis of Expressions of Biomarkers and Their Prognostic Value in Head and Neck Squamous Cell Carcinoma

### 2021-06-27 Cover_letter_JPM.tex
Answers done (draft0

[materials and methods]

%RefSeq Release 38 (November 7, 2009)

%[Dataset]
The Gene Expression Omnibus (GEO) database, founded by \acrfull{ncbi}, is a public repository supporting MIAME-compliant data submissions of microarray and sequence-based experiments.
R package of GEOquery\cite{Sean2007} was used to
download RNA-Seq dataset (in SOFT or MINiML format) of a HNSCC cohort, GSE65858, from the GEO database (available at https://www.ncbi.nlm.nih.gov/geo/geo2r/?\\acc=GSE65858). 
GSE65858\cite{Wichmann2015} has OS event, RFS event and survival time.
There are 270 HNSCC participants involved in this cohort.
%The number of genes (or probes of RNA-Seq in platform GPL10558, Illumina HumanHT-12 V4.0 expression beadchip) is 30,330.
%n=270 n= 273 = ?? + 168(alive) + 88(dead) = 256
The expression data was generated using the platform GPL10558 (Illumina HumanHT-12 v4.0 Expression BeadChip), which targets on more than 30,330 annotated genes (47,000 probes, derived from the \acrshort{ncbi} Reference Sequence, release 38 on November 7, 2009). % (RefSeq) 
GSE65858 is a dataset for candidate selection in our workflow.
We have also performed Kaplan-Meier and Cox survival analyses, under the cutoff of each gene expression at their median value.\\

[Results]
TCGA HNSCC cohort has been applied for exploration of the biomarker candidate
Total 9416 genes has %\acrshort{fdr}-
unadjusted \textit{P} value less than 0.05.
The 967 genes were kept by criteria of uncorrected \textit{P} value (below 0.05) and \acrfull{hr} (greater than 1.8 or less than 0.6).
%The 10 genes, named as DKK1, CAMK2N1, STC2, PGK1, SURF4, USP10, NDFIP1, FOXA2, STIP1, and DKC1, are significantly associated with the poor prognosis of overall survival (OS). 
%The other ten genes are over-expressed in the better survival group (with hazard ratio $<=0.6$), named as ZNF557, ZNF266, IL19, MYO1H, FCGBP, LOC148709, EVPLL, PNMA5, IQCN (previous name as KIAA1683), and NPB.
Furthermore, CAMK2N1, IL19, FCGBP and the other 17 genes (DKK1, STC2, PGK1, SURF4, USP10, NDFIP1, FOXA2, STIP1, DKC1, ZNF557, ZNF266, MYO1H, LOC148709, EVPLL, PNMA5, IQCN (previous name as KIAA1683), and NPB) have significant FDR-adjusted \textit{P} values ($<0.0003$) in Kaplan-Meier estimate, and greater hazard ratio (HR) ($> 1.8$ or $< 0.6$) in Cox's model (please see Figure \ref{fig:hazards3}; $log_{10}(0.0003)=-3.5$).
%During our exploration phase with TCGA cohort, the 20 genes, listed
%six overexpressed genes (symbol as CAMK2N1, PGK1, SURF4, USP10, NDFIP1, FOXA2) are bad guy;
%four overexpressed genes (symbol as IL19, FCGBP, IQCN - former symbol as KIAA1683, and NPB) are good guy.
%}
%in Figure \ref{fig:hazards3}, have FDR-adjusted \textit{P} values ($<0.001$) in Kaplan-Meier estimate, and survival impact on HR ($> 1.8$ or $< 0.6$) in Cox's model.

@@

[Discussion]
[more in the discussion section]
1) % (ok)
Statistical significance (\textit{P} value) is affected by both sample size, error, effect size (substantive significance)\cite{Sullivan2012}\cite{Thiese2016}, and cutoff. 
The effect size is the magnitude of the difference (or relationship) among comparing groups.
%A \textit{P} value is a tool to help interpret findings from biomedical research.
%In reporting and interpreting studies, both 
%the substantive significance (effect size) and statistical significance (P value) are essential results to be reported.
While statistical significance can inform the researcher whether an effect exists, 
the \textit{P} value will not directly tell the effect size.
%The effect size is the main finding of a quantitative study.
%Statistical significance depends upon both sample size and effect size. 
The effect size is independent of the sample size\cite{Sullivan2012}.

@@

%%%%

2) in TCGA HNSCC cohort, 
cutoff at optimal by sliding-window or just median:
when unadjusted \textit{P} value $<0.05$, the gene number: 6284 vs. 3118, 967 vs. 209.
% HNSCC_OS_marginS_THREE_pvalue005_noCancerGene
Result by optimal cutoff: total 9416 plots,
6284 are raw KM P value < 0.05
534?? genes are FDR KM P value <0.05
967 genes which HR is greater than 1.5 or less than 0.5.
Result by median cutoff: total 20080 genes, 
3118 are raw KM P value < 0.05; 
209 genes are FDR KM P value <0.05
*** IL19 is excluded by this way :-) (special S-curve P value plot)
%差別很大，張大網
%([2021/06/27] run code: if median cut for TCGA KM Cox by *** pvalueTex code)
discussion about sliding-window pvalueTex to find the more potential candidate, while small effect size could be double-checked by GSE65858.
(  )2-3 put in supplementary figure %Rplot_pvaluePlot_NDFIP1.pdf
%列入 workflow in materials and methods;

% ****** IL19 必須要靠 pvalueTex 才能發現
\textit{P} value plot with a acute S-curve bending on the median zone, it lets the FDR-KM \textit{P} value has large difference between median cutoff (\num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.1154315}) and optimal cutoff (\num{6.54e-6}).
A sliding-window selection of cutoffs could boost their statistic significance to pass the correction by even FDR or Bonferroni methods.
%> gene_labeled_golden[!gene_labeled_golden %in% fdr100_KMsurvival$Gene.Symbol[fdr100_KMsurvival$FDR.Pvalue<0.05]]
[1] "IL19"
***NDFIP1
the special cases of gene, such as NDFIP1, DKC1, PNMA5, and NPB % ; % 給他們一個機會
(ok) Answer 2-1 %加入 discussion
( )  Answer 2-3 supplementary  %圖就好 figure Rplot_pvaluePlot_NDFIP1.pdf

For example, a gene NDFIP1, with a \textit{P} value of 0.05 at 50\% quantile cutoff, could achieve a \textit{P} value of \num{1e-6} at 70\% quantile cutoff.
NDFIP1 has the FDR-corrected \textit{P} value as \num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.0001074233}.
However, it fails as the candidate by less effect size (HR = 1.33 in GSE65858 cohort, comparing with HR = 2.027 in TCGA cohort).
% (ok) Answer 2-1 加入 discussion
According to the results of sliding-window cutoff finding, the preliminary biomarkers have FDR-corrected \textit{P} values from \num{1e-4} to 0.04 which have been grouped by cutoffs between 50\% and 70\% quantile (Please see Figure \ref{fig:figure4}; it is also shown in manuscript as updated "Figure 4": top 3 candidate biomarkers, CAMK2N1, IL19, and FCGBP).
Moreover, the gene NDFIP1, one of our 20 preliminary candidates, has a \textit{P} value (around $0.05$) at 50\% quantile cutoff, achieving a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff (please see Figure \ref{fig:pvaluePlot_NDFIP1}, and also in 
supplementary figure S1). After the FDR correction, NDFIP1 still has \textit{P} value of \num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.0001074233}.
%Rplot_pvaluePlot_NDFIP1.pdf
%  to find cases_OS outside of range 180-220 (== cutoff not at median)
%# pvalue plot as 倒V字形 "^" or W p_OS pvalues plots (with local minials)
% # 99 genes have one p_OS >= 0.05, in OS_pvalueTex005
%# 3641 genes have one p_OS >= 0.02, in OS_pvalueTex002
%# special case: HNSCC_survivalAnalysis_marginS_NDFIP1.xlsx
%# NDFIP1 (within our 20 preliminary candidates), 
In those \textit{P}-value plots, we could see that the sliding-window selection of cutoff finding might collect more potential candidates (\textit{P} values far less than $0.001$) for further Cox's modeling.
%Some useful biomarkers might be kick-out prematurely in "median cutoff" of survival analysis.
Indeed, it needs to be validated by the other independent cohort (e.x. GSE65858).

>>>>

[materials and methods]\\
%RefSeq Release 38 (November 7, 2009) 
%[Dataset]
The Gene Expression Omnibus (GEO) database, founded by \acrfull{ncbi}, is a public repository supporting MIAME-compliant data submissions of microarray and sequence-based experiments.
R package of GEOquery\cite{Sean2007} was used to
download RNA-Seq dataset (in SOFT or MINiML format) of a HNSCC cohort, GSE65858, from the GEO database (available at https://www.ncbi.nlm.nih.gov/geo/geo2r/?\\acc=GSE65858). 
GSE65858\cite{Wichmann2015} has OS event, RFS event and survival time.
There are 270 HNSCC participants involved in this cohort.
%The number of genes (or probes of RNA-Seq in platform GPL10558, Illumina HumanHT-12 V4.0 expression beadchip) is 30,330.
%n=270 n= 273 = ?? + 168(alive) + 88(dead) = 256
The expression data was generated using the platform GPL10558 (Illumina HumanHT-12 v4.0 Expression BeadChip), which targets on more than 30,330 annotated genes (47,000 probes, derived from the \acrshort{ncbi} Reference Sequence, release 38 on November 7, 2009). % (RefSeq) 
GSE65858 is a dataset for candidate selection in our workflow.
We have also performed Kaplan-Meier and Cox survival analyses, under the cutoff of each gene expression at their median value.\\

[Results]\\
TCGA HNSCC cohort has been applied for exploration of the biomarker candidate
Total 9416 genes has %\acrshort{fdr}-
unadjusted \textit{P} value less than 0.05.
The 967 genes were kept by criteria of uncorrected \textit{P} value (below 0.05) and \acrfull{hr} (greater than 1.8 or less than 0.6).
%The 10 genes, named as DKK1, CAMK2N1, STC2, PGK1, SURF4, USP10, NDFIP1, FOXA2, STIP1, and DKC1, are significantly associated with the poor prognosis of overall survival (OS). 
%The other ten genes are over-expressed in the better survival group (with hazard ratio $<=0.6$), named as ZNF557, ZNF266, IL19, MYO1H, FCGBP, LOC148709, EVPLL, PNMA5, IQCN (previous name as KIAA1683), and NPB.
Furthermore, CAMK2N1, IL19, FCGBP and the other 17 genes (DKK1, STC2, PGK1, SURF4, USP10, NDFIP1, FOXA2, STIP1, DKC1, ZNF557, ZNF266, MYO1H, LOC148709, EVPLL, PNMA5, IQCN (previous name as KIAA1683), and NPB) have significant FDR-adjusted \textit{P} values ($<0.0003$) in Kaplan-Meier estimate, and greater hazard ratio (HR) ($> 1.8$ or $< 0.6$) in Cox's model (please see Figure \ref{fig:hazards3}; $log_{10}(0.0003)=-3.5$).
%During our exploration phase with TCGA cohort, the 20 genes, listed
%six overexpressed genes (symbol as CAMK2N1, PGK1, SURF4, USP10, NDFIP1, FOXA2) are bad guy;
%four overexpressed genes (symbol as IL19, FCGBP, IQCN - former symbol as KIAA1683, and NPB) are good guy.
%}
%in Figure \ref{fig:hazards3}, have FDR-adjusted \textit{P} values ($<0.001$) in Kaplan-Meier estimate, and survival impact on HR ($> 1.8$ or $< 0.6$) in Cox's model.

@@

[Discussion]\\
%[more in the discussion section]
1) \\% (ok)
Statistical significance (\textit{P} value) is affected by both sample size, error, effect size (substantive significance)\cite{Sullivan2012}\cite{Thiese2016}, and cutoff. 
The effect size is the magnitude of the difference (or relationship) among comparing groups.
%A \textit{P} value is a tool to help interpret findings from biomedical research.
%In reporting and interpreting studies, both 
%the substantive significance (effect size) and statistical significance (P value) are essential results to be reported.
While statistical significance can inform the researcher whether an effect exists, 
the \textit{P} value will not directly tell the effect size.
%The effect size is the main finding of a quantitative study.
%Statistical significance depends upon both sample size and effect size. 
The effect size is independent of the sample size\cite{Sullivan2012}.

@@

%%%%
%(ok)2-3 put in supplementary figure S1 %Rplot_pvaluePlot_NDFIP1.pdf
%from Answer 2-3 supplementary Figure S1  %圖就好 figure Rplot_pvaluePlot_NDFIP1.pdf

2) \\
( ) % pvalueTex is essential
There is benefit of using sliding-window cutoff method in survival analysis of TCGA HNSCC cohort.
We have compared the results of cutoff at optimal by sliding-window or just at median of gene expression.
It shows
the gene number with unadjusted \textit{P} value $<0.05$, which is 6284 vs. 3118, respectively.
After FDR-correction, it shows 967 vs. 209, respectively.
% HNSCC_OS_marginS_THREE_pvalue005_noCancerGene
%Result by optimal cutoff: total 9416 plots,
%6284 are raw KM P value < 0.05
%534?? genes are FDR KM P value <0.05
%967 genes which HR is greater than 1.5 or less than 0.5.
%Result by median cutoff: total 20080 genes, 
%3118 are raw KM P value < 0.05; 
%209 genes are FDR KM P value <0.05
The sliding-window cutoff method could catch more potential candidates (\textit{P} values far less than $0.001$) for Cox's modeling at next stage in our workflow.
These preliminary candidates must be validated by Bonferroni correction and the other independent cohorts (GSE65858, GSE2837) to prevent the false discovery.

%small effect size could be double-checked by GSE65858.
%差別很大，張大網
%([2021/06/27] run code: if median cut for TCGA KM Cox by *** pvalueTex code)
%列入 workflow in materials and methods;

We will explain above mentioned by two examples.
%***NDFIP1  % 給他們一個機會
Reviewing the special cases of gene, such as NDFIP1, DKC1, PNMA5, and NPB,
we have noticed that NDFIP1, with a \textit{P} value of 0.05 at 50\% quantile cutoff, could achieve a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff.
NDFIP1 has the FDR-corrected \textit{P} value as \num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.0001074233} (please see supplementary figure S1). % it is \ref{fig:pvaluePlot_NDFIP1
%Moreover, the gene NDFIP1, one of our 20 preliminary candidates, has a \textit{P} value (around $0.05$) at 50\% quantile cutoff, achieving a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff (please see Figure \ref{fig:pvaluePlot_NDFIP1}, and also in supplementary figure S1).
%After the FDR correction, NDFIP1 still has \textit{P} value of \num[round-precision=3, round-mode=figures, scientific-notation=true]{0.0001074233}.
% the effect size
Although it has failed to compete as a candidate by less effect size (HR = 1.33) in GSE65858 cohort, compared with one (HR = 2.027) in TCGA cohort%.
, and the other (HR = 1.42) in GSE2837 cohort.
% (ok) Answer 2-1 加入 discussion
%According to the results of sliding-window cutoff finding, the preliminary biomarkers have FDR-corrected \textit{P} values from \num{1e-4} to 0.04 which have been grouped by cutoffs between 50\% and 70\% quantile (Please see Figure \ref{fig:figure4}; it is also shown in manuscript as updated "Figure 4": top 3 candidate biomarkers, CAMK2N1, IL19, and FCGBP).

%Rplot_pvaluePlot_NDFIP1.pdf
%  to find cases_OS outside of range 180-220 (== cutoff not at median)
%# pvalue plot as 倒V字形 "^" or W p_OS pvalues plots (with local minials)
% # 99 genes have one p_OS >= 0.05, in OS_pvalueTex005
%# 3641 genes have one p_OS >= 0.02, in OS_pvalueTex002
%# special case: HNSCC_survivalAnalysis_marginS_NDFIP1.xlsx
%# NDFIP1 (within our 20 preliminary candidates), 
%In those \textit{P}-value plots, we could see that the sliding-window selection of cutoff finding might collect more potential candidates (\textit{P} values far less than $0.001$) for further Cox's modeling.
%Some useful biomarkers might be kick-out prematurely in "median cutoff" of survival analysis.
%Indeed, it needs to be validated by the other independent cohort (e.x. GSE65858).

% ****** IL19 必須要靠 pvalueTex 才能發現
%*** IL19 is excluded by this way :-) (special S-curve P value plot)
The other example is one of our candidate: IL19.
It has a \textit{P} value plot with acute S-curve bending at the median zone, that lets the FDR-KM \textit{P} value has large difference between median cutoff (\num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.1154315}) and optimal cutoff (\num{6.54e-6}).
The sliding-window selection of cutoffs could boost their statistic significance to pass the correction by both FDR and Bonferroni methods.
%> gene_labeled_golden[!gene_labeled_golden %in% fdr100_KMsurvival$Gene.Symbol[fdr100_KMsurvival$FDR.Pvalue<0.05]]
%[1] "IL19"
Fortunately, IL19 has passed to become a candidate by better effect size (HR = 0.472 in TCGA cohort), and also by the validation with the other independent cohort (HR = 0.630 in GSE65858 cohort%).
, and HR = 0.84 in GSE2837).

% ***using GSE2837 or not??? ***
%(x)
%After the discovery by our workflow, we validated the those 20 genes in the other independent \acrshort{hnscc} cohort (GSE2837).
%The survival significance (Kaplan-Meier \textit{P} value) is in the following probes among 10 genes:\\
%=> Quantile normalization
%1) poor prognosis: CAMK2N1 (0.048214), PGK1 (0.009978), SURF4 (0.023127), USP10 (0.017768), NDFIP1 (0.022758, HR= 1.42), FOXA2 (0.001587);\\ % FOXA2 (0.038125)
%=> no normalization
%2) better prognosis: IL19 (0.049731, HR= 0.84), FCGBP (0.005658), KIAA1683 (IQCN, 0.005886), NPB (0.014177);\\
